### 2017-03-16 Manuscripts/cog-psych/cogpsych.tex
intro drafted

Theoretical accounts of these effects were primarily grounded within the categorization literature. For example, the predominant ``Path of Least Resistance'' account \citep[see][]{ward1994structured,ward1995s,ward2002role} proposes that, when generating a new species of animal, people retrieve from memory a known subcategory of animals (e.g., \textit{bird}, \textit{dog}, \textit{horse}), and simply change some of the features to make something new. People were thought to change only features that are not characteristic of the retrieved category (e.g., if \textit{bird} was retrieved, the presence of \textit{wings} would not change, but \textit{color} might). This theory incorporates elements of the highly influential basic-level categories framework \citep{rosch1975cognitive,rosch1976basic}, as well as the exemplar view \citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. Problematically, however, the experimental paradigms employed in these studies were relatively uncontrolled, precluding the development of formal approaches. Specifically, participants in these studies are typically asked to generate objects that they have a rich degree of prior knowledge about (e.g., plants, animals, tools, toys), and so it is difficult to apply formal models to the data.

@@

The authors evaluated the predictions of several formal models on their data. Most notably, they showed that a Bayesian hierarchical sampling model provided the strongest account. Their model represents categories as multivariate normal distributions within the space, and from exposure to members of observed categories the model induces a domain-wide representation of the distributional commonalities among the observed categories. When generating a new category, the model samples a new distribution from the domain, and thus the distributional properties of observed categories will be preserved within the generates category.

@@

It is worth noting that the literature reviewed above provides a somewhat limited view of category generation: the main concern of the published experiments has been on the distributional correspondences between learned and generated categories, and as a result most of the modeling efforts have been towards explaining those effects. In this paper, we seek to identify other important constraints on the processes involved in creative generation. Specifically, we focus on the constraints imposed by category contrast. That is, in order to generate a novel concept, individuals must produce something that is in some capacity \textit{different} from what they already know. Thus, in a trivial sense, contrast can be viewed as a fundamental constraint on creative generation: new concepts must firstly be different from existing ones.

Although it is evident that people are \textit{capable} of creating new concepts and categories, it is not altogether clear how new concepts are systematically made different from what is already known. The hierarchical sampling model developed by \cite{jern2013probabilistic}, for example, does not provide any strong claims about how generated categories should contrast with learned ones: it is simply assumed that categories are generated in a random location in the domain.

>>>>

Theoretical accounts of these effects have primarily been grounded within the categorization literature. For example, the predominant ``Path of Least Resistance'' account \citep[see][]{ward1994structured,ward1995s,ward2002role} proposes that, when generating a new species of animal, people retrieve from memory a known subcategory of animals (e.g., \textit{bird}, \textit{dog}, \textit{horse}), and simply change some of the features to make something new. People were thought to change only features that are not characteristic of the retrieved category (e.g., if \textit{bird} was retrieved, the presence of \textit{wings} would not change, but \textit{color} might). This theory incorporates elements of the highly influential basic-level categories framework \citep{rosch1975cognitive,rosch1976basic}, as well as the exemplar view \citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. Problematically, however, the experimental paradigms employed in these studies were relatively uncontrolled, precluding the development of formal approaches. Specifically, participants in these studies are typically asked to generate objects that they have a rich degree of prior knowledge about (e.g., plants, animals, tools, toys), and so it is difficult to apply formal models to the data.

@@

The authors evaluated the predictions of several formal models on their data. Most notably, they showed that a Bayesian hierarchical sampling model provided the strongest account. Their model represents categories as multivariate normal distributions within the space, and from exposure to category members the model induces a domain-wide representation of the distributional commonalities among the observed categories. When generating a new category, the model samples a new distribution from the domain, and thus the distributional properties of observed categories will be preserved within the generates category.

@@

It is worth noting that the literature reviewed above provides a somewhat limited view of category generation: the main concern of the published experiments has been on the distributional correspondences between learned and generated categories, and as a result most of the modeling efforts have been towards explaining those effects. In this paper, we investigate another important constraint on the creative use of conceptual knowledge: category contrast. In order to generate a novel concept, individuals must produce something that is in some capacity \textit{different} from what they already know. Thus, in a trivial sense, contrast can be viewed as a fundamental constraint on creative generation: new concepts must firstly be different from existing ones.

Although it is evident that people are \textit{capable} of creating new concepts and categories, it is not entirely clear how new concepts are systematically made different from what is already known. The hierarchical sampling model developed by \cite{jern2013probabilistic}, for example, does not provide any strong claims about how generated categories should contrast with learned ones: the model only proposes that generated categories are sampled from the same underlying domain distribution as observed categories, and will thus share a common distributional structure. The authors do not make predictions about the \textit{location} of the category within the domain (the perceptual instantiation of category members). Indeed, under a strict interpretation of the hierarchical Bayesian approach, the generated category should actually overlap substantially with the known categories (as it's location is drawn based on those categories). Alternatively, the model's placement of the new category may simply be random (this was the case in the simulations conducted by \citeauthor{jern2013probabilistic}).

\nbcnote{Maybe tune that language to be more in line with the probabilistic world?}

The copy-and-tweak model tested by \cite{jern2013probabilistic} also claims little about how generated categories should contrast with what is already known. In their simulations, the model was only tested on generation after the learner had been exposed to members of the target category, and so the model's ability to generate a new category was not evaluated. However, the model's generation is based exclusively on similarity to known members of the target category; when there are no members of the target category, generation is presumably random.

\nbcnote{Maybe mention something about path of least resistance prediction that contrast is on non-characteristic dimensions? }

\subsection{Overview}

As noted in the sections above, research on the creative generation of novel concepts has thusfar focused on the finding that generated categories tend to possess distributional commonalities with known categories. However, a fundamental goal of concept generation is to create something \textit{new} (i.e., different from what is already known). The manner in which generated categories differ from known ones is, nonetheless, poorly understood: existing formal approaches do not make strong predictions about how creatively generated concepts should systematically differ from existing ones.

In the sections below, we present two experiments demonstrating systematic effects of category contrast on creative generation. We also introduce a novel, exemplar-based model to explain these effects: PACKER (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}). Finally, we report formal simulations comparing PACKER to the hierarchical sampling and copy-and-tweak models developed by \cite{jern2013probabilistic}.
